Reviving stochasticity: SMBH-binary eccentricity is inherently random

### 2023-06-12 papers/paper-eccentricity/paper-eccentricity.tex

\begin{figure*}
    \centering
    \includegraphics{sample_model_orbits}
    \caption{
    Top:
    Sample orbits computed with the toy model for potentials with different degrees of asymmetry
    given by the spheroid eccentricity $e_\mathrm{s}$, with otherwise identical initial conditions.
    The dashed lines show the isopotential contours of the model.
    Only the early part of the evolution is shown for clarity.
    Bottom:
    The orbital eccentricity $e$ of the model orbits.
    The vertical line marks the end of the period shown in the top panels. 
    Note the non-linear scale to allow showing the high-eccentricities reached by
    the $e_\mathrm{s}=0.2$ model.
    }
    \label{fig:model_orbits}
\end{figure*}

To gain insight into the origin of the dependence of binary eccentricity on the 
deflection angle $\_\mathrm{defl}$ during the binary formation,
we have constructed a simple toy model that can qualitatively reproduce the observed
behavior from the simulations.
This model includes only the essential components relevant to this process,
which are the asymmetric background stellar potential,
dynamical friction from the stellar background,
and naturally the SMBHs themselves.

For simplicity, we consider only symmetrical 1:1 mergers, and take the BH orbits
to lie in the $xy$-plane.
In this case the relative motion of the BHs can be treated as a one-body problem,
with the equation of motion
\begin{equation} \label{eq:toy_model_eom}
\ddot{\vb{x}} = -\frac{2 G M_\bullet}{|\vb{x}|^3} \vb{x} + \vb{a}_\mathrm{bg} + \vb{a}_\mathrm{DF},
\end{equation}
where $\vb{x}=(x,y)$ is the separation vector of the BHs, $M_\bullet$ the mass of a single BH,
$\vb{a}_\mathrm{bg}$ the acceleration due to the background potential,
and $\vb{a}_\mathrm{DF}$ the acceleration due to dynamical friction.

The background galaxy is modeled as a constant density spheroidal potential,
causing the acceleration \citep[e.g.][]{binney2008}
\begin{equation}
\vb{a}_\mathrm{bg} = - 2 \pi G \rho \operatorname{diag}(A_x,A_y) \vb{x},
\end{equation}
where $\rho$ is the stellar density,
and the $A$ coefficients are related to the eccentricity $e_\mathrm{s}$ of the spheroid as
\begin{align}
A_x &= 2 \frac{1-e_\mathrm{s}^2}{e_\mathrm{s}^2} 
    \left[\frac{1}{2 e_\mathrm{s}} \ln\left(\frac{1+e_\mathrm{s}}{1-e_\mathrm{s}}\right) - 1 \right]\\
A_y &= \frac{1-e_\mathrm{s}^2}{e_\mathrm{s}^2} 
        \left[ \frac{1}{1-e_\mathrm{s}^2} 
        - \frac{1}{2 e_\mathrm{s}} \ln\left(\frac{1+e_\mathrm{s}}{1-e_\mathrm{s}}\right)\right].
\end{align}
Here we have chosen the long axis of the spheroid to be aligned with the $x$-axis.
With eccentricity values of $e_\mathrm{s} \approx 0.9$,
this spheroidal potential matches the central potential in the galaxy during the
SMBH binary binding phase reasonably well,
apart from the contributions from the BHs themselves and the groups of stars bound to them.

For dynamical friction, we use the \citet{chandrasekhar1943} formula assuming
a Maxwellian distribution with a constant velocity dispersion $\sigma_\mathrm{M}$ \citep[e.g.][]{binney2008}.
This formula gives the acceleration of a single BH due to dynamical friction as
\begin{equation}\label{eq:chandra_df}
\begin{aligned}
\vb{a}_{1,\mathrm{DF}} = &-\frac{4 \pi G^2 M_\bullet \rho \ln{\Lambda}}{|\vb{v}_{1}|^3}\\
                        &\times\left(\operatorname{erf}(X) - \frac{2X}{\sqrt{\pi}} \exp(-X^2)\right)\vb{v}_{1},
\end{aligned}
\end{equation}
where $\vb{v}_1 = \dot{\vb{x}}/2$ is the velocity of a single BH,
$X = |\vb{v}_1|/\sqrt{2}\sigma_\mathrm{M}$,
and the Coulomb logarithm is taken to have the value $\ln{\Lambda} = 10$.
To account for differences between the constant density background of stars
assumed here and the more complex system that is present in the simulations
we additionally multiply the acceleration given by Eq.~\eqref{eq:chandra_df}
with a scaling and cut-off function
\begin{equation}
f(a) = \frac{\alpha}{1 + \exp[(a_\mathrm{c}-a)/d_\mathrm{c}]},
\end{equation}
which serves to remove the effect of dynamical friction when the SMBH binary has become
hard in addition to allowing the strength of dynamical friction to be scaled to
achieve similar behavior as observed in the simulations, since in the simulations the
stellar background is much more complex.
Here $a$ is the semi-major axis of a bound binary, and for unbound binaries we set $f(a)=\alpha$.
We set the scaling factor $\alpha=0.5$, and use cut-off scales $a_\mathrm{c}=2 a_\mathrm{h}$
and $d_\mathrm{c} = 0.5 a_\mathrm{h}$,
with $a_\mathrm{h}$ being the semimajor-axis of a hard BH binary.
The total dynamical friction term is then
\begin{equation}
\vb{a}_{\mathrm{DF}} = 2 f(a) \vb{a}_{1,\mathrm{DF}}.
\end{equation}

To mimic the nearly linearly plunging orbit in the high-eccentricity merger orbit simulations,
we specify the initial conditions confined to the $xy$-plane as 
\begin{align}
\vb{x}_0 &= (r_0, b)\\
\dot{\vb{x}}_0 &= (-v_0, 0).
\end{align}
We fix $r_0 = 25\,\mathrm{pc}$,
and use model parameters
$\rho = 300 \,\Msun\,\mathrm{pc}^{-3}$,
$\sigma_\mathrm{M}=200\,\mathrm{km\,s^{-1}}$,
and
$M_\bullet=10^8\,\Msun$,
which are fairly similar to the central properties of the merging galaxy when
the SMBHs are becoming bound in the $e_0=0.90$ and $e_0=0.99$ merger orbit simulations.
Based on the simulations we set $v_0 = 450\,\mathrm{km\,s^{-1}}$ to match the $e_0=0.90$ case,
and $v_0 = 560\,\mathrm{km\,s^{-1}}$ to match the $e_0=0.99$ case.
The spheroid eccentricity parameter is not easily measurable from the simulations,
and we therefore perform calculations with different values of $e_\mathrm{s}$.
We compute the resulting eccentricities for a range of impact parameters up to $b=20\,\mathrm{pc}$
for the two initial velocities,
by solving the equation of motion \eqref{eq:toy_model_eom} until the binary has become hard using
the error controlled 8th order Runge--Kutta method DOP853 included in the SciPy library \citep{virtanen2020}.
The range of impact parameters covers most of the relevant parameter space,
as a $90\degr$ deflection during the first pericenter passage
corresponds to initial impact parameters of $b\approx 5.8\,\mathrm{pc}$ and
$b\approx 3.3\,\mathrm{pc}$ for the $e_0=0.90$ and $e_0=0.99$ cases, respectively.

The resulting model curves are shown together with the simulation data in \autoref{fig:theta_e}.
We observe a modest decrease in deflection angle $\theta_\mathrm{defl}$ with increasing mass resolution for both the $e_0=0.90$ and $e_0=0.99$ cases. 
To correctly match the data in the $e_0=0.90$ case, the model curve has been shifted to the left by $12\degr$,
while in the $e_0=0.99$ case a smaller shift of $6\degr$ is used.
Otherwise the model appears to give a very good match to the simulation results,
and the required slight shift could be due to the galactic potential being slightly tilted in the simulations 
compared to the BH trajectories due to the complex dynamics at this stage.
In the $e_0=0.99$ case we can also see that the simulation data has significant scatter around
the model curve in the $\theta_\mathrm{defl} \gtrsim 120\degr$ region,
although the model does appear to capture the mean behavior fairly well.
The behavior of the model in this region is also quite sensitive to the exact value of $e_\mathrm{s}$,
which might be related to the large scatter in the simulation data.

Overall we can however see that this simple model correctly captures the behavior of eccentricity seen in the simulations.
We can therefore use the model to understand the cause of the complicated dependence of the binary eccentricity
on the deflection angle $\theta_\mathrm{defl}$.
\autoref{fig:model_orbits} shows sets of orbits computed with the model for two different potential shapes,
with the parameters otherwise as in the model that matches the $e_0=0.90$ simulation results.
In the $e_\mathrm{s}=0.9$ case the shape of the orbit after the initial scattering depends strongly on the deflection angle,
as the asymmetric potential causes torque on the SMBHs that are not moving along the main axes of the potential.
On the other hand, in the more symmetric potential with $e_\mathrm{s}=0.2$,
The orbit shape is not very sensitive to the deflection angle due to the symmetry of the system,
and the binary always reaches extremely high eccentricities.

>>>>

\begin{figure*}
    \centering
    \includegraphics{sample_model_orbits}
    \caption{
    Top:
    Sample orbits computed with the toy model for different background potential eccentricities $e_\mathrm{s}$,
    with otherwise identical initial conditions.
    The dashed lines show the isopotential contours of the model.
    Only the early part of the evolution is shown for clarity.
    Bottom:
    The orbital eccentricity $e$ of the model orbits (note the non-linear scale).
    The vertical line marks the end of the period shown in the top panels. 
    }
    \label{fig:model_orbits}
\end{figure*}

The dependency of the binary eccentricity on the 
deflection angle $\theta_\mathrm{defl}$ during the binary formation
can be reproduced using a simple toy model,
which includes only the essential components relevant to this process.
Taking the orbit to lie in the $xy$-plane,
the relative motion of the two equal-mass SMBHs in this model is follows the 
equation of motion
\begin{equation} \label{eq:toy_model_eom}
\ddot{\vb{x}} = -\frac{2 G M_\bullet}{|\vb{x}|^3} \vb{x} + \vb{a}_\mathrm{bg} + \vb{a}_\mathrm{DF},
\end{equation}
where $\vb{x}=(x,y)$ is the separation vector of the SMBHs,
$M_\bullet = 10^8\,\Msun$ is the mass of a single SMBH,
$\vb{a}_\mathrm{bg}$ the acceleration due to the asymmetric background potential,
and $\vb{a}_\mathrm{DF}$ the acceleration due to dynamical friction.

The stellar background is modeled as a constant density spheroidal potential
\citep[e.g.][]{binney2008}
\begin{equation}
\Phi_\mathrm{bg}(\vb{x}) = \pi G \rho (A_x x^2 + A_y y^2),
\end{equation}
with the acceleration given by 
\begin{equation}
\vb{a}_\mathrm{bg}(\vb{x}) = - \nabla \Phi_\mathrm{bg}(\vb{x}).
\end{equation}
The $A$ coefficients are related to the eccentricity $e_\mathrm{s}$ of the spheroid as
\begin{align}
A_x &= 2 \frac{1-e_\mathrm{s}^2}{e_\mathrm{s}^2} 
    \left[\frac{1}{2 e_\mathrm{s}} \ln\left(\frac{1+e_\mathrm{s}}{1-e_\mathrm{s}}\right) - 1 \right]\\
A_y &= \frac{1-e_\mathrm{s}^2}{e_\mathrm{s}^2} 
        \left[ \frac{1}{1-e_\mathrm{s}^2} 
        - \frac{1}{2 e_\mathrm{s}} \ln\left(\frac{1+e_\mathrm{s}}{1-e_\mathrm{s}}\right)\right],
\end{align}
with the long axis aligned along the $x$-axis.
The stellar density is set to $\rho = 300 \,\Msun\,\mathrm{pc}^{-3}$,
which approximately matches the values seen in the simulations when SMBH binary is becoming bound.

Dynamical friction is modeled using the \citet{chandrasekhar1943} formula assuming
a Maxwellian distribution with a constant velocity dispersion
${\sigma_\mathrm{M}=200\,\mathrm{km\,s^{-1}}}$,
set based on the value measured from the simulations.
This formula gives the acceleration of a single BH due to dynamical friction as
\citep[e.g.][]{binney2008}
\begin{equation}\label{eq:chandra_df}
\begin{aligned}
\vb{a}_{1,\mathrm{DF}} = &-\frac{4 \pi G^2 M_\bullet \rho \ln{\Lambda}}{|\vb{v}_{1}|^3}\\
                        &\times\left(\operatorname{erf}(X) - \frac{2X}{\sqrt{\pi}} \exp(-X^2)\right)\vb{v}_{1}.
\end{aligned}
\end{equation}
Here $\vb{v}_1 = \dot{\vb{x}}/2$ is the velocity of a single BH, 
$X = |\vb{v}_1|/\sqrt{2}\sigma_\mathrm{M}$,
and the Coulomb logarithm is taken to have the value $\ln{\Lambda} = 5$,
which is appropriate for the size of the simulated systems.

The effect of dynamical friction weakens as the SMBH binary orbit shrinks.
To account for this, when the binary is bound 
we multiply the acceleration given by Eq.~\eqref{eq:chandra_df}
with the smooth cut-off function
\begin{equation}
f(a) = \frac{1}{1 + \exp[(a_\mathrm{c}-a)/d_\mathrm{c}]}.
\end{equation}
Here $a$ is the semi-major axis of a bound binary,
and the cut-off scales are $a_\mathrm{c}=2 a_\mathrm{h}$
and $d_\mathrm{c} = 0.5 a_\mathrm{h}$,
with $a_\mathrm{h}$ defined by equation \eqref{eq:ahard}.
The total dynamical friction term is then
\begin{equation}
\vb{a}_{\mathrm{DF}} = 2 f(a) \vb{a}_{1,\mathrm{DF}}.
\end{equation}

To mimic the nearly linearly plunging orbits before the scattering event seen in \autoref{fig:orbit},
we specify the initial conditions as 
\begin{align}
\vb{x}_0 &= (25\,\mathrm{pc}, b)\\
\dot{\vb{x}}_0 &= (-v_0, 0).
\end{align}
Based on the simulations, we set $v_0 = 450\,\mathrm{km\,s^{-1}}$ to match the
typical SMBH relative velocity at this separation in the $e_0=0.90$ case,
and $v_0 = 560\,\mathrm{km\,s^{-1}}$ to match the $e_0=0.99$ case.
The spheroid eccentricity parameter is not easily measurable from the simulations,
due to stellar components that remain bound to the SMBHs and obscure the background
potential relevant for the dynamics,
as well as due to the potential in the simulations being constantly evolving.
We therefore perform calculations with different values of $e_\mathrm{s}$,
finding a good fit to simulation results for $e_\mathrm{s}\approx 0.9$.

We compute the resulting eccentricities for a range of impact parameters up to $b=20\,\mathrm{pc}$
for the two initial velocities,
by solving the equation of motion \eqref{eq:toy_model_eom} until the binary has become hard using
the error controlled 8th order Runge--Kutta method DOP853 included in the SciPy library \citep{virtanen2020}.
For reference, a $90\degr$ deflection during the first pericenter passage
corresponds to initial impact parameters of $b\approx 5.8\,\mathrm{pc}$ and
$b\approx 3.3\,\mathrm{pc}$ for the $e_0=0.90$ and $e_0=0.99$ cases, respectively.

The resulting model curves are shown together with the simulation data in \autoref{fig:theta_e}.
To correctly match the data in the $e_0=0.90$ case, the model curve has been shifted to the left by $12\degr$,
while in the $e_0=0.99$ case a smaller shift of $6\degr$ is used.
These shifts are likely required due to the rotation of the stellar component tilting the background potential
relative to the SMBH trajectories in the simulations, with the effect being stronger in the $e_0=0.90$ case
due to the less radial merger orbit.
In the $e_0=0.99$ case we can also see that the simulation data has significant scatter around
the model curve in the $\theta_\mathrm{defl} \gtrsim 120\degr$ region,
although the model does appear to capture the mean behavior fairly well.
The behavior of the model in this region is also quite sensitive to the exact value of $e_\mathrm{s}$,
which might be related to the large scatter in the simulation data.

Overall we can however see that this simple model correctly captures the behavior of eccentricity seen in the simulations.
We can therefore use the model to understand the cause of the complicated dependence of the binary eccentricity
on the deflection angle $\theta_\mathrm{defl}$.
\autoref{fig:model_orbits} shows sets of orbits computed with the model for two different potential shapes,
with the parameters otherwise as in the model that matches the $e_0=0.90$ simulation results.
In the $e_\mathrm{s}=0.9$ case the shape of the orbit after the initial scattering depends strongly on the deflection angle,
as the asymmetric potential causes torque on the SMBHs that are not moving along the main axes of the potential.
On the other hand, in the more symmetric potential with $e_\mathrm{s}=0.2$,
the orbit shape is not very sensitive to the deflection angle due to the symmetry of the system,
and the binary always reaches extremely high eccentricities.
